---
title: "Part 2 - Basic Inferential Data Analysis"
author: "Prafful Agrawal"
date: "July 7, 2020"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
```

## Overview

In this project, we undertake a basic *inferential* data analysis to investigate the **`ToothGrowth`** data in R's **`datasets`** package. The **`ToothGrowth`** dataset consists of data on the effect of Vitamin C on the tooth growth in Guinea Pigs. It has details of **60** guinea pigs where each animal received one of the *three* dose levels of Vitamin C (**0.5**, **1** and **2** mg/day) by one of the *two* delivery methods, *orange juice* or *ascorbic acid* (coded as **OJ** and **VC** respectively).


## Data Preprocessing

```{r packages}
library(ggplot2)
library(dplyr)
library(tidyr)
```

Load the **`Toothgrowth`** data and look at the first few rows.

```{r data}
library(datasets)
data("ToothGrowth")

# Look at first few rows of data
head(ToothGrowth)
```

Look at the structure of the dataset.

```{r structure}
str(ToothGrowth)
```

Convert **`dose`** to a **factor** variable.

```{r factor}
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
str(ToothGrowth)
```

## Basic Exploratory Data Analysis

Construct a plot showing the variation of tooth growth with respect to the *dose* and *supplement delivery method*.

```{r Plot-1, cache=TRUE, fig.height=4, fig.width=7}
g <- ToothGrowth %>%
      ggplot(aes(x = dose, y = len, fill = supp)) +
        geom_boxplot() +
        labs(title = "Variation of tooth growth with respect to dose and supplement delivery method",
             x = "Dose (in mg/day)",
             y = "Tooth growth",
             fill = "Supplement")

print(g)
```

From the above graph, we can observe that the tooth growth increases as the *dose* increases. Also, for low and medium *doses* (**0.5** and **1** mg/day), the *supplement delivery method* **OJ** shows a higher tooth growth than the *supplement delivery method* **VC**, whereas the tooth growth is nearly the same for high *dose* (**2** mg/day) irrespective of the *supplement delivery method*.

Let us compute the means of the tooth growth length with respect to the *dose* and *supplement delivery method*.

```{r pivot}
Tooth <- ToothGrowth %>%
          group_by(supp, dose) %>%
          summarize(mean = mean(len)) %>%
          pivot_wider(names_from = dose, values_from = mean,
                      names_prefix = "dose ")

print(Tooth)
```

This is in accordance with our previous observations.


## Statistical Inference Analysis

Let us now test if our previous observations are statistically significant.

Since, the number of observations *n* are small, we will perform **One-sided T-test** to analyse the effect of *dose* and *supplement delivery method* on the tooth growth.

We will assume that the samples are **representative** of the population and the variance of the different samples are **different**. Also, the samples are **independent** from each other.

### Variation of tooth growth with respect to dose

Our *Null* hypothesis states that the difference in the means of tooth growth with respect to *dose* is **Equal to Zero** while the *Alternative* hypothesis states that it is **Greater than Zero**, i.e.

$$H_{0}: \Delta_{Mean}^{1} = 0$$

$$H_{a}: \Delta_{Mean}^{1} > 0$$

We will consider only the *high* and *low* *doses* (**2** and **0.5** mg/day respectively).

```{r t-test 01}
g1 <- ToothGrowth$len[ToothGrowth$dose == '0.5']
g2 <- ToothGrowth$len[ToothGrowth$dose == '2']

t1 <- t.test(g2 - g1, paired = FALSE, var.equal = FALSE, alternative = "greater")
```

The 95% confidence interval and the p-value are:

```{r confidence interval 01}
print(paste("Confidence interval:",
            paste(round(t1$conf.int, 2),
                  collapse = " - ")))
```

```{r p-value 01}
print(paste0("p-value: ", signif(t1$p.value, 3)))
```

Thus, the *Null* hypothesis can be **rejected**, i.e. there is an increase in tooth growth with respect to increasing *dose*.

### Variation of tooth growth with respect to supplement delivery method

Our *Null* hypothesis states that the difference in the means of tooth growth with respect to *supplement delivery method* is **Equal to Zero** while the *Alternative* hypothesis states that it is **Greater than Zero**, i.e.

$$H_{0}: \Delta_{Mean}^{2} = 0$$


$$H_{a}: \Delta_{Mean}^{2} > 0$$

```{r t-test 02}
g3 <- ToothGrowth$len[ToothGrowth$supp == 'VC']
g4 <- ToothGrowth$len[ToothGrowth$supp == 'OJ']

t2 <- t.test(g4 - g3, paired = FALSE, var.equal = FALSE, alternative = "greater")
```

The 95% confidence interval and the p-value are:

```{r confidence interval 02}
print(paste("Confidence interval:",
            paste(round(t2$conf.int, 2),
                  collapse = " - ")))
```

```{r p-value 02}
print(paste0("p-value: ", signif(t2$p.value, 3)))
```

Thus, the *Null* hypothesis can be **rejected**, i.e. there is an increase in tooth growth with the use of *supplement delivery method* **OJ** over *supplement delivery method* **VC**.


## Conclusion

1. Increasing the *dose* of the Vitamin C results in an increase of tooth growth.
2. Using **OJ** as *supplement delivery method* results in a higher tooth growth over using **VC** as *supplement delivery method*.


## Possible shortcomings in the analysis

Since, there are two variables affecting the tooth growth simultaneously, the analysis will perform better for **paired** data.


## Appendix

Load the packages.

```{r, echo=TRUE, eval=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
```

Load the data.

```{r, echo=TRUE, eval=FALSE}
library(datasets)
data("ToothGrowth")

# Look at first few rows of data
head(ToothGrowth)
```

Structure of the dataset.

```{r, echo=TRUE, eval=FALSE}
str(ToothGrowth)
```

Converting **`dose`** to a **factor** variable.

```{r, echo=TRUE, eval=FALSE}
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
str(ToothGrowth)
```

Code for Exploratory plot.

```{r, echo=TRUE, eval=FALSE}
g <- ToothGrowth %>%
      ggplot(aes(x = dose, y = len, fill = supp)) +
        geom_boxplot() +
        labs(title = "Variation of tooth growth with respect to dose and supplement delivery method",
             x = "Dose (in mg/day)",
             y = "Tooth growth",
             fill = "Supplement")

print(g)
```

Code for Pivot tables of the means.

```{r, echo=TRUE, eval=FALSE}
Tooth <- ToothGrowth %>%
          group_by(supp, dose) %>%
          summarize(mean = mean(len)) %>%
          pivot_wider(names_from = dose, values_from = mean,
                      names_prefix = "dose ")

print(Tooth)
```

Code for T-test 01.

```{r, echo=TRUE, eval=FALSE}
g1 <- ToothGrowth$len[ToothGrowth$dose == '0.5']
g2 <- ToothGrowth$len[ToothGrowth$dose == '2']

t1 <- t.test(g2 - g1, paired = FALSE, var.equal = FALSE, alternative = "greater")
```

```{r, echo=TRUE, eval=FALSE}
print(paste("Confidence interval:",
            paste(round(t1$conf.int, 2),
                  collapse = " - ")))
```

```{r, echo=TRUE, eval=FALSE}
print(paste0("p-value: ", signif(t1$p.value, 3)))
```

Code for T-test 02.

```{r, echo=TRUE, eval=FALSE}
g3 <- ToothGrowth$len[ToothGrowth$supp == 'VC']
g4 <- ToothGrowth$len[ToothGrowth$supp == 'OJ']

t2 <- t.test(g4 - g3, paired = FALSE, var.equal = FALSE, alternative = "greater")
```

```{r, echo=TRUE, eval=FALSE}
print(paste("Confidence interval:",
            paste(round(t2$conf.int, 2),
                  collapse = " - ")))
```

```{r, echo=TRUE, eval=FALSE}
print(paste0("p-value: ", signif(t2$p.value, 3)))
```